---
title: "Exercise Chapter 1"
author: "Tobias Hugentobler"
date: "22.02.2023"
output:
  html_document:
    toc: yes
    toc_float: yes
    df_print: paged
    theme: paper
    code_folding: show
    math_method: katex
---
```{r include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, 
  warning = FALSE,
  message = FALSE, 
  fig.height = 5,
  fig.width = 6,
  eval = TRUE
)
```

## Exercise 1

Load the data `dataCar` from the package "insuranceData". It represents claim data on vehicle insurance policies from 2004 to 2005. Some variables like "gender" describe the policy holder, others like "veh_age" the vehicle, and some variables carry information on claims, e.g. "numclaims". Each row represents policy information valid in a certain time window. Use the pipe, "dplyr", and "ggplot2" to solve the following tasks.

```{r}
library(tidyverse)
library(insuranceData)
library(plotly)

data(dataCar)

str(dataCar)
head(dataCar)
```

### a)

Draw barplots of the discrete variables "numclaims", "agecat" (categorized driver age), and "gender".

```{r}

  dataCar %>% ggplot(mapping = aes(x = numclaims)) +
    geom_bar(fill = "navyblue")

  dataCar %>% ggplot(mapping = aes(x = agecat)) +
    geom_bar(fill = "navyblue")
    
  dataCar %>% ggplot(mapping = aes(x = gender)) +
    geom_bar(fill = "navyblue")
```

### b)

Draw a histogram of the vehicle value "veh_value" (in 10'000 Australian Dollars). Truncate values above 7 (this means: if a value is larger than 7, set it to 7).

```{r}
dataCar %>% mutate(veh_value = (veh_value > 7)*7 + veh_value*(veh_value <= 7)) %>% 
  arrange(-veh_value) %>% 
  ggplot(mapping = aes(veh_value)) +
  geom_histogram(fill = "navyblue")

#alternative
dataCar %>%
  mutate(veh_value = pmin(7, veh_value)) %>% #pmin is rowwise_min but much more efficient 
  ggplot(aes(veh_value)) +
  geom_histogram(fill = "orange") +
  labs(
    title = "Histogram of vehicle Values",
    x = "Behicle value[10'000] AUD"
  )
```
### c)

Calculate the average number of claims per level of "agecat" and visualize the result as a scatterplot. Interpret the result.

```{r}
dataCar %>%
  group_by(agecat) %>% 
  summarize(avg_claims = mean(numclaims)) %>% 
  ggplot(mapping = aes(x = agecat, y = avg_claims)) +
  geom_point(fill = "navy")
```

The older the owner the smaller the average claim gets. That makes sense, since younger driver may drive more reckless than older people.

### d)

Bin "veh_value" into quartiles and analyze its association with the number of claims as in 1c.

```{r}
summary(dataCar$veh_value)
```


```{r}
plot <- dataCar %>% 
  mutate(veh_value_bin = ntile(veh_value, n=4)) %>% 
  group_by(veh_value_bin) %>% 
  summarize(avg_claims = mean(numclaims)) %>% 
  ggplot(mapping = aes(x = veh_value_bin, y = avg_claims)) +
  geom_point(fill = "navy")


plot
```

The higher the price of the car, the higher are the average claims.

### e)

Use the "plotly" package to turn the plot from d. interactive.

```{r}
(plot) %>% ggplotly()
```

## Exercise 3

The sieve of Eratosthenes is an ancient algorithm to get all prime numbers up to any given limit $n$, see [Wikipedia](https://en.wikipedia.org/wiki/Sieve_of_Eratosthenes). Write a function `sieve_of_eratosthenes(n)` that returns all prime numbers up to $n$. Benchmark the results for $n = 10^5$  with the package "microbenchmark". Mind your coding style!

```{r}

sieve_of_eratostheses <- function(n){
  sieve = !logical(n)
  i = 2
  while(i <= sqrt(n)){
    if(sieve[i]){
      j = i^2
      while(j<=n){
        sieve[j]=FALSE
        j = j+i
      }
    }
    i = i+1
  }
  out = which(sieve %in% TRUE)
  out = out[out!=1]
  return(out)
}

```

```{r}
library(microbenchmark)

res <- microbenchmark(sieve_of_eratostheses(10^5), times=100)
```


```{r}
print(res)
ggplot2::autoplot(res)
```

## Exercise 4

In Exercise 1c, we have calculated and plotted the average number of claims per level of "agecat" in the `dataCar` data.
    a. Write a function `avg_claim_counts(v)` that provides such a visualization for any discrete variable `v`.
    b. Extend this function with a second argument `interactive` to control whether the resulting plot is interactive or not.

```{r}
avg_claim_counts <- function(v, interactive=FALSE){
  plot <- dataCar %>%
    group_by(across(all_of(v))) %>% 
    summarize(avg_claims = mean(numclaims)) %>% 
    ggplot(mapping = aes(x = .data[[v]], y = avg_claims)) +
    geom_point(fill = "navy")
  
  if(interactive){
    plot <- (plot) %>% ggplotly()
  }
  
  return(plot)
}
```

## Exercise 5

Extend the “student” class from Section “plot, print, summary” by the optional information “semester”. It represents the number of semesters the student is already registered. Add a summary() method that would neatly print the name and the semester of the student.

```{r}
student <- function(given_name, family_name, semester = NULL) {
  out <- list(
    given_name = given_name,
    family_name = family_name,
    semester = semester
  )
  class(out) <- "student"
  out
}

summary.student <- function(object){
  cat("Name: ", object$given_name, " ", object$family_name, "\n")
  cat("Semester: ", object$semester, "\n")
}

#other option to set a Method for a class

# setMethod("summary", "student", function(object) {
#   cat("Name: ", object$given_name, " ", object$family_name, "\n")
#   cat("Semester: ", object$semester, "\n")
# })

me <- student("Tobias", "Hugentobler", 2)
summary(me)
```


